---
title: "Representative Population"
author: "Belen Zapata-Diomedi, Steve Pemberton"
email: "mbzd2@cam.ac.uk, mb2592@cam.ac.uk, steve.pemberton@rmit.edu.au"
date: today
format: 
  html:
    code-fold: true
    embed-resources: true
    df-print: paged
    toc: true
    number-sections: true
execute: 
  echo: false
  message: false
  warning: false
editor_options:
  mode: visual
  markdown:
    wrap: 72
---

# Introduction

This document [generates a representative population of 10,100 people to
whom the health model will be applied].

```{r setup, include=FALSE}

rm(list = ls())

library(tidyverse)  # Includes dplyr, tidyr, readr, ggplot2, purrr, etc.
library(knitr)      # Report generation
library(gt)         # Pretty tables
library(here)       # File path management
library(haven)      # Read .dta files
library(Hmisc)      # Quantiles

# Set root directory one level up from current directory
knitr::opts_knit$set(root.dir = dirname(getwd()))

# Avoid scientific notation
options("scipen"=100, "digits"=4)

```

# Mortality deciles

A life table for England is  created in `health_data.Rmd`, based on (1) mortality statistics by age and sex, and (2) relative risks by lower layer super output areas (LSOAs), aggregated by local authority districts (LADs) and index of multiple deprivation (IMD).

The following plots show deciles of mortality rates by age and sex, calculated using the LAD/IMD lifetable.

```{r}

# load life tables

england_lifetable_lad_imd <- read.csv(here("health_data/processed/england_lifetable_lad_imd.csv"))

# calculate deciles, weighted by population, picking the middle of each decile

deciles <- england_lifetable_lad_imd %>%
  group_by(sex, age) %>%
  summarise(
    decile_5 = wtd.quantile(rate1000, weights = population, probs = 0.05, na.rm = TRUE),
    decile_15 = wtd.quantile(rate1000, weights = population, probs = 0.15, na.rm = TRUE),
    decile_25 = wtd.quantile(rate1000, weights = population, probs = 0.25, na.rm = TRUE),
    decile_35 = wtd.quantile(rate1000, weights = population, probs = 0.35, na.rm = TRUE),
    decile_45 = wtd.quantile(rate1000, weights = population, probs = 0.45, na.rm = TRUE),
    decile_55 = wtd.quantile(rate1000, weights = population, probs = 0.55, na.rm = TRUE),
    decile_65 = wtd.quantile(rate1000, weights = population, probs = 0.65, na.rm = TRUE),
    decile_75 = wtd.quantile(rate1000, weights = population, probs = 0.75, na.rm = TRUE),
    decile_85 = wtd.quantile(rate1000, weights = population, probs = 0.85, na.rm = TRUE),
    decile_95 = wtd.quantile(rate1000, weights = population, probs = 0.95, na.rm = TRUE)) %>%
  ungroup()
      

# plot preparation
deciles_long <- deciles %>%
  pivot_longer(
    cols = starts_with("decile_"),
    names_to = "decile",
    values_to = "rate1000"
  ) %>%
  mutate(decile = factor(decile, 
                         levels = paste0("decile_", c(5,15,25,35,45,55,65,75,85,95)),
                         labels = paste0("Decile ", c(5,15,25,35,45,55,65,75,85,95))))

p_deciles <- ggplot(deciles_long, aes(x = age, y = rate1000, color = decile)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ sex) +
  scale_color_brewer(palette = "RdYlBu", name = "Decile") +  # good for ordinal data
  labs(
    x = "Age (years)",
    y = "Mortality rate (per 1000)",
    title = "Weighted deciles of mortality rate by age and sex"
  ) +
  scale_y_log10() +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom"
  )

print(p_deciles)

```


# mMETs distribution

The synthetic population of 3 million people, with their physical activity details (walking, cycling and sport), is used to calculate Marginal Metabolic Equivalent of Tasks (mMET) values.

The following plots show the distribution of the mMET values by age and sex.

```{r}

# load synthetic population
sp <- read_dta(here("synthetic_pop/SPindivid_CensusNTSALS.dta"))

# calculate mMETs for synthetic population
# see https://github.com/StevePem/that-brisbane/blob/master/Scripts/data_prep/mmet_pp.R for values
MMET_WALKING <- 2.5
MMET_CYCLING <- 5.8
MMET_VIG <- 7


sp_mmets <- sp %>%
  # assume 0 sport_wkhr for under age 5 (shown as NA)
  mutate(sport_wkhr = ifelse(sport_wkhr < 0 | is.na(sport_wkhr), 0, sport_wkhr)) %>%
  # calculate mmets
  mutate(mmets = walktime_wkhr * MMET_WALKING + cycletime_wkhr * MMET_CYCLING + sport_wkhr * MMET_VIG,
         sex = as.factor(ifelse(female == 1, "Females", "Males"))) %>%
  # keep required fields
  dplyr::select(census_id, sex, age1year, mmets)

# plot the densities
# 1. Bin ages into 10-year groups and factorize
sp_mmets_binned <- sp_mmets %>%
  filter(!is.na(mmets)) %>%
  mutate(
    age_group = cut(
      age1year,
      breaks = c(seq(0, 80, by = 10), 90),
      labels = c("0-9","10-19","20-29","30-39","40-49","50-59","60-69","70-79","80-89"),
      include.lowest = TRUE,
      right = FALSE
    ),
    age_group = factor(age_group, levels = c("0-9","10-19","20-29","30-39","40-49","50-59","60-69","70-79","80-89"))
  )

# 2. Calculate percentiles per sex
percentiles <- sp_mmets_binned %>%
  group_by(sex) %>%
  summarise(
    p25  = quantile(mmets, 0.25, na.rm = TRUE),
    p50  = quantile(mmets, 0.50, na.rm = TRUE),
    p75  = quantile(mmets, 0.75, na.rm = TRUE),
    p95  = quantile(mmets, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = starts_with("p"), names_to = "percentile", values_to = "mmets") %>%
  mutate(label = case_when(
    percentile == "p25" ~ "25th",
    percentile == "p50" ~ "50th",
    percentile == "p75" ~ "75th",
    percentile == "p95" ~ "95th"
  ))

# 3. Compute max density per sex for placing labels
max_density_per_sex <- sp_mmets_binned %>%
  group_by(sex) %>%
  do({
    dens <- density(.$mmets, adjust = 1.5)
    data.frame(max_density = max(dens$y))
  }) %>%
  distinct()

# 4. Plot densities with log1p x-axis
p_mmets_percentiles <- ggplot(sp_mmets_binned, 
                              aes(x = mmets, y = after_stat(density), color = age_group, group = age_group)) +
  geom_density(alpha = 0.4, linewidth = 0.7, adjust = 1.5) +
  facet_wrap(~ sex) +
  scale_color_viridis_d(option = "C", 
                        name = "Age group (years)",
                        guide = guide_legend(nrow = 2, byrow = TRUE)) +
  scale_x_continuous(trans = "log1p",  
                     breaks = c(0, 1, 10, 50),
                     labels = c(0, 1, 10, 50)) +
  labs(
    x = "MMETs (hrs/week, log1p scale)",
    y = "Density",
    title = "Distribution of MMETs by 10-year Age Group and Sex"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.major.x = element_blank(),  # remove vertical gridlines
    panel.grid.minor.x = element_blank()
  ) +
  # Add percentile lines without legend
  geom_vline(data = percentiles, aes(xintercept = mmets), 
             color = "grey40", linewidth = 0.7, linetype = "dashed") +
  # Add percentile labels above the densities
  geom_text(data = percentiles,
            aes(x = mmets, y = 0.9, label = label),
            inherit.aes = FALSE,
            angle = 90, vjust = -0.5, hjust = 0,
            size = 3,
            color = "grey20")

print(p_mmets_percentiles)

```

# Representative population

The representative population is [10,100 people: 101 ageyears * 2 genders * 10 mortality risk groups * 5 mMETs groups.]

The mortality groups are the middle of the decile values; the mMETs groups are the middle of the quintile values.

```{r}
# mMETs quintiles
mmets_percentiles <- sp_mmets %>%  # SP note - as above, this is altered to convert NA sport_wkhr  (for age < 5) to zeros
  rename(age = age1year) %>%
  
  # calculate quintiles
  group_by(sex, age) %>%
  summarise(quintile_10 = quantile(mmets, probs = 0.1, na.rm = TRUE),
            quintile_30 = quantile(mmets, probs = 0.3, na.rm = TRUE),
            quintile_50 = quantile(mmets, probs = 0.5, na.rm = TRUE),
            quintile_70 = quantile(mmets, probs = 0.7, na.rm = TRUE),
            quintile_90 = quantile(mmets, probs = 0.9, na.rm = TRUE)) %>%
  ungroup() %>%
  
  # expand the age sequence to include 90–100 for each sex
  complete(sex, age = 0:100) %>%
  
  # assume mMETs is zero for ave > 90
  mutate(across(starts_with("quintile"),
                ~ if_else(age > 89, 0, .x))) %>%

  # pivot longer to produce one row per qunitile
  pivot_longer(
    cols = starts_with("quintile_"),
    names_to = "mmets_quintile",
    values_to = "mmets"
  ) %>%
  mutate(mmets_quintile = factor(mmets_quintile, 
                         levels = paste0("quintile_", c(10,30,50,70,90)))) 

# representative population using LAD deciles
rep_pop <- england_lifetable_lad_imd %>%
  
  # calculate deciles [or, as above, maybe should be 5th, 15th etc percentiles]
  group_by(sex, age) %>%
  summarise(
    decile_5 = wtd.quantile(rate1000, weights = population, probs = 0.05, na.rm = TRUE),
    decile_15 = wtd.quantile(rate1000, weights = population, probs = 0.15, na.rm = TRUE),
    decile_25 = wtd.quantile(rate1000, weights = population, probs = 0.25, na.rm = TRUE),
    decile_35 = wtd.quantile(rate1000, weights = population, probs = 0.35, na.rm = TRUE),
    decile_45 = wtd.quantile(rate1000, weights = population, probs = 0.45, na.rm = TRUE),
    decile_55 = wtd.quantile(rate1000, weights = population, probs = 0.55, na.rm = TRUE),
    decile_65 = wtd.quantile(rate1000, weights = population, probs = 0.65, na.rm = TRUE),
    decile_75 = wtd.quantile(rate1000, weights = population, probs = 0.75, na.rm = TRUE),
    decile_85 = wtd.quantile(rate1000, weights = population, probs = 0.85, na.rm = TRUE),
    decile_95 = wtd.quantile(rate1000, weights = population, probs = 0.95, na.rm = TRUE)) %>%
  ungroup() %>%
  
  # pivot longer to produce one row per decile
  pivot_longer(
    cols = starts_with("decile_"),
    names_to = "mortality_decile",
    values_to = "rate1000"
  ) %>%
  mutate(mortality_decile = factor(mortality_decile, 
                                   levels = paste0("decile_", c(5,15,25,35,45,55,65,75,85,95)))) %>%
  
  # add mMETs quintiles 
  left_join(mmets_percentiles, by = c("sex", "age"), relationship = "many-to-many")


# save output
write.csv(rep_pop, here("health_data/processed/rep_pop.csv"), row.names= FALSE)


```